### 2018-01-27 submission/manuscript.Rmd
Look over edits after incorporating Ada suggestions

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness after correcting for study effect and variable region sequenced with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. There are also genera like *Roseburia* that some studies find to be increased in CRC but other studies report either decreased or no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue directly may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

@@

***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
Next we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. The top 5 most significantly increased RR were the same between adenoma and carcinoma for both stool and colon tissue [Figure 3]. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

@@

In stool, the most common genera in the top 10 most imporant variables, in models for the first approach, that used full genera-based models were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B]. Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B]. For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

@@

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
The results that we have presented show that the both genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why the these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.

@@

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study. Relative risk significance testing utilized the chi-squared test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

>>>>

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in CRC but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

@@

***Lower Bacterial Diversity is Associated with Increased RR of CRC:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

@@

In stool, the most common genera in the top 10 most imporant variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B]. Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B]. For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

@@

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
The results that we have presented show that both the genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.

@@

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study. Relative risk significance testing utilized the chi-squared test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].
